### 2017-09-13 chapters/introduction/physiology.tex
Update to latest version

Deoxgenated blood is transported from the body trough the veins and to
the right atrium (RA). From the RA, blood flows to the right ventricle
(RV), before it is sent through the the pulmonary arteries(PA) and to the
lungs, where the bloods gets oxygenated. From the lungs, the blood
travels through the pulmonary veins to the left atrium (LA), and
finally to the left ventricle (LV), before is is sent out to the body
again through the aorta.

@@

The pressure and volume inside the ventricle can provide us with
information about the passive and active properties of the
myocardium. Imaging it was possible to shut down all contractile
units in the myocardium, and remove all loads (e.g set the pressure
to zero). This would be analogous to have a deflated balloon. In order to
get some information about the stiffness of the myocardium one could
try to inflate the ventricle. The stiffer the myocardium is, the
higher pressure would be needed in order to increase the volume.
This relationship between the pressure and volume during the inactive
state is called the end diastolic pressure volume relationship
(\emph{EDPVR}).

Similarly, imaging now that the ventricle is at the end-systolic state,
when the muscle cells are contracting at its most forcefully. Changing
the pressure at this state would also result in a change in volume and
this relationship between pressure and volume  is called the end
systolic pressure volume relationship (\emph{ESPVR}). The two pressure
volume relationships are depicted in Figure \ref{fig:intro_pvr}.

@@

\subsubsection{Heart failure}
Heart failure is a common term for all heart diseases in which the
heart in unable to supply the body with enough blood to meet its demand.
% Consequently, patients suffering from heart failure has a reduced
% cardiac output. It is common to distinguish between systolic and
% diastolic heart failure.

@@

One type of heart failure can result from conduction block in the
heart. The action potential travels through these electrical highways
of high conducting cells. One of these highways known as the Bundle of
His splits into two parts, the right and left bundle. The right bundle
activates the right ventricle, while the left bundle activates the
left ventricle. For patients suffering from left bundle branch block,
there is ``road block'' on the highway along the left bundle, meaning
that the left ventricle is activated later than preferable, and the
right ventricle is activated prior to the left
ventricle. This can result in a dyssynchronous contraction, yielding a
reduced pumping effect. A promising treatment for this patients is
called Cardiac Resynchronization Therapy (CRT) \cite{cleland2005effect}.

@@

\subsubsection{Pulmonary hypertension}

\subsection{Summary of Cardiac Physiology}
The multiscale and multiphysics phenomena governing the mechanics of
the heart is complex. How the molecular dynamics should be coupled to
the electrophysiology at the cellular level, how the electrophysiology
should be coupled to the mechanics at the organ level and what type of
feedback mechanisms that should be included is far from fully
understood. In this thesis we limit ourselves to the study of what is
happening on the organ level, and concentrates purely on the
mechanics.

>>>>

Deoxgenated blood is transported from the body through the veins and to
the right atrium (RA). From the RA, blood flows to the right ventricle
(RV), before it is sent through the the pulmonary arteries(PA) and to the
lungs, where the bloods gets oxygenated. From the lungs, the blood
travels through the pulmonary veins to the left atrium (LA), and
finally to the left ventricle (LV), before is is sent out to the body
again through the aorta.

@@

The pressure and volume inside the ventricle can provide us with
information about the passive and active properties of the
myocardium. Imagine it was possible to shut down all contractile
units in the myocardium, and remove all loads (e.g set the pressure
to zero). This would be analogous to have a deflated balloon. In order to
get some information about the stiffness of the myocardium one could
try to inflate the ventricle. The stiffer the myocardium is, the
higher pressure would be needed in order to increase the volume.
This relationship between the pressure and volume during the inactive
state is called the end diastolic pressure volume relationship
(\emph{EDPVR}).

Similarly, imagine now that the ventricle is at the end-systolic state,
when the muscle cells are contracting at its most forcefully. Changing
the pressure at this state would also result in a change in volume and
this relationship between pressure and volume  is called the end
systolic pressure volume relationship (\emph{ESPVR}). The two pressure
volume relationships are depicted in Figure \ref{fig:intro_pvr}.

@@

\subsubsection{Heart failure}
Heart failure (HF) is a common term for all heart diseases in which the
heart is unable to supply the body with enough blood to meet its
demand. It is common to separate between diastolic and systolic HF.
Diastolic HF refers to the heart's inability to be passively
filled, and thereby reduce the cardiac output according to the 
Frank-Starling law of the heart. In systolic HF on the other had, the
heart's ability to contract efficiently is reduced.

% Consequently, patients suffering from heart failure has a reduced
% cardiac output. It is common to distinguish between systolic and
% diastolic heart failure.

@@

One type of heart failure can result from conduction block in the
heart. The action potential travels through these electrical highways
of high conducting cells. One of these highways known as the Bundle of
His splits into two parts, the right and left bundle. The right bundle
activates the right ventricle, while the left bundle activates the
left ventricle. For patients suffering from left bundle branch block,
there is ``road block'' on the highway along the left bundle, meaning
that the left ventricle is activated later than preferable, and the
right ventricle is activated prior to the left
ventricle. This can result in a dyssynchronous contraction, yielding a
reduced pumping effect. A promising treatment for these patients is
called Cardiac Resynchronization Therapy (CRT) \cite{cleland2005effect}.

@@

% \subsubsection{Pulmonary hypertension}

\subsection{Summary of Cardiac Physiology}
The multiscale and multiphysics phenomena governing the mechanics of
the heart is complex. How the molecular dynamics should be coupled to
the electrophysiology at the cellular level, how the electrophysiology
should be coupled to the mechanics at the organ level and what type of
feedback mechanisms that should be included is far from fully
understood. In this thesis we limit ourselves to the study of what is
happening on the organ level, and concentrate purely on the mechanical
aspect.

### 2017-09-13 chapters/introduction/summary.tex
Update to latest version

\section{Summary of papers}
In this final introductory section we summarize the papers that make
up this thesis, and we also give some final concluding remarks
and future directions.

\subsection{Paper 1: High-resolution data assimilation of cardiac mechanics
  applied to a dyssynchronous ventricle}
In this paper we develop and test the pipeline for constructing a
patient specific mechanical simulation of a patient's heart, based on
clinical measurments and adjoint-based data assimilation
techniques.

@@

The pipeline is divided into two phases, a passive phase where we
estimate the linear isotropic parameters $a$ in
\eqref{eq:holzapel_trans} as a global material parameter using the
measurement points belonging atrial systole, and a active phase where we estimate the
a spatially varying contraction parameter ($\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}) at each measurement point
with active contraction. During the passive phase we only fit the
volumes, while during the active phase 51 strain measurements in the radial,
longitudinal and circumferential direction in each AHA segment (Figure
\ref{fig:echopac_output}) are used in the optimization.

@@

\subsection{Paper 2: Estimating cardiac contraction through high resolution
  data assimilation of a personalized mechanical model}
In this paper we apply the method developed in the previous paper to a
cohort of patients and estimate indices of cardiac contractility. More
specifically, a group of seven patients diagnosed with left bundle branch block and
selected for cardiac resynchronization therapy (CRT) and a group of
seven healthy control subjects where included in the study.

@@

\subsection{Paper 3: Mechanical Analysis of pulmonary hypertension via
  adjoint based data assimilation of a finite element model.}

@@

A rule of thump is that \emph{the spatial resolution of the parameters should be
  reflected in the spatial resolution of the observations}. This means
that if one is trying to fit data that are spatially resolved at some level,
then choosing parameters that are resolved at a finer level should be
done with caution. Regarding both paper 1 and 2 we see that the
spatial resolution chosen was at a much finer level than the input
data. In this case, regularization techniques was used to restrict the
parameter space.

When estimating high dimensional parameters, a natural question
concerning uniqueness of these estimates arises. It is obvious that if one allows for
enough degree of freedom in the parameter space, then it is possible
to fit almost any type of data. Therefore \emph{more work on ensuring
identifiability} of these estimates (Section
\ref{sec:intro_identifiability}) should be done. If the output of such
model should have any clinical utility, then uniqueness of the
estimated parameters absolutely pivotal. Moreover, \emph{validation} of
these model is what matter most in terms of translating such computational
models into the clinic.

Regarding the mechanical modeling of the heart, 
\emph{choosing appropriate boundary conditions} that reflect the reality has
been a problem during the work in this thesis, and several different
choices has been made. Moreover, \emph{accounting for the orthotropic as
well as the visco-elastic behavior of the myocardium} is something that should
be investigated in future studies. Accounting for an orthotropic
behavior would acquire more parameters to be estimated, which would
thus require more input data.

@@

The choice of active model for the myocardium is another topic that
should be addressed in more detail in future studies. In the author's
opinion, the output of the two fundamentally different approaches may
vary a lot. In particular, it is evident that the active strain
formulation do a better job in fitting strain data. One
hypothesis is that the amount of transverse active stresses should be
adjusted to each individual. These transverse active stresses are
naturally embedded in the active strain approach via a volume
preserving active deformation.

Finally, coupling the electrophysiology to the mechanical model within
this adjoint based data assimilation framework is something that
should also be investigated in the future.

% Here we start by listing a couple of statements
% which should be taken into considerations.

>>>>

\section{Summary of papers}
\label{sec:summary}
In this final introductory section we summarize the papers that make
up this thesis, and we also give some final concluding remarks
and future directions.

\subsection{Paper 1: High-resolution data assimilation of cardiac mechanics
  applied to a dyssynchronous ventricle}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.8\textwidth]{chapters/introduction/figures/paper1}
\caption{Summary figure for paper 1}
\label{fig:paper1}
\end{figure}

In this paper we develop and test the pipeline for constructing a
patient specific mechanical simulation of a patient's heart, based on
clinical measurments and adjoint-based data assimilation
techniques.

@@

The pipeline is divided into two phases, a passive phase where we
estimate the linear isotropic parameters $a$ in
\eqref{eq:holzapel_trans} as a global material parameter using the
measurement points belonging atrial systole, and an active phase where we estimate the
a spatially varying contraction parameter ($\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}) at each measurement point
with active contraction. During the passive phase we only fit the
volumes, while during the active phase 51 strain measurements in the radial,
longitudinal and circumferential direction in each AHA segment (Figure
\ref{fig:echopac_output}) are used in the optimization.

@@

\subsection{Paper 2: Estimating cardiac contraction through high resolution
  data assimilation of a personalized mechanical model}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.8\textwidth]{chapters/introduction/figures/paper2}
\caption{Summary figure for paper 2}
\label{fig:paper2}
\end{figure}

In this paper we apply the method developed in the previous paper to a
cohort of patients and estimate indices of cardiac contractility. More
specifically, a group of seven patients diagnosed with left bundle branch block and
selected for cardiac resynchronization therapy (CRT) and a group of
seven healthy control subjects where included in the study.

@@

\subsection{Paper 3: Personalized bi-ventricular mechanical analysis
  using gradient-based optimization..}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.8\textwidth]{chapters/introduction/figures/paper3}
\caption{Summary figure for paper 3}
\label{fig:paper3}
\end{figure}

In this paper we extend the work in the two previous paper to
bi-ventricular geomtries, and use it to estimate patient-specific
myofiber stress and indices of contractility. Furthermore, we investigate
the senstivtiy of these computed features the helical fiber angle and
the choice of active modeling framework.

In this work we estimate the linear isotopic material parameter in
\eqref{eq:holzapel_trans}, spatially resolved on the LV and RV, by
minimizing the error in end-diastolic LV and RV cavity volumes. We also
apply a simplified algorithm for estimating in unloaded geometry, in
order to overcome issues related to estimating an unloaded
BiV-geometry, which is not well-posed. The amount of active
contraction is estimated during the active phase, by minimizing
simulated and meausured volumes and circumferential strain. The active
control parameter, which are $\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}  and $T_a$ in
\eqref{eq:intro_active_stress} for the active strain and active stress
formulation respectively, are spatially resolved on the
LV free wall (LVFW), the RV free wall (RVFW) and the septum. With so
few control parameters, issues related to conflicting objectives, i.e
that is is not possible to minimize both the strain and volume, is
evident. However, with a low dimensional parameter space, questions
regarding identifiability and uniqueness of the estimated parameters
are easier to answer. In this study we also perform a validation of
the model, by comparing simulated and measured longitudinal strain
which is not used in the optimization.

The results show low variability with respect to choice of fiber angle
and active modeling framework. Also, the fit of data depends upon the
choice of fiber angle, and a steeper fiber angle than previously
suggested, provides the best fit. The work here could potentially be
used to extract patient-specific maps of ventricular fiber stress and
contractility which could be useful in diagnostic of several heart
diseases related to heart failure.

@@

A rule of thump is that \emph{the spatial resolution of the parameters should be
  reflected in the spatial resolution of the observations}. This means
that if one is trying to fit data that are spatially resolved at some level,
then choosing parameters that are resolved at a finer level should be
done with caution. Regarding both paper 1, 2 and 4 we see that the
spatial resolution chosen was at a much finer level than the input
data. In this case, regularization techniques was used to restrict the
parameter space.

Regarding the mechanical modeling of the heart, 
\emph{choosing appropriate boundary conditions} that reflect the reality has
been an issue during the work of this thesis, and several different
choices has been made. Moreover, \emph{accounting for the orthotropic as
well as the visco-elastic behavior of the myocardium} is something that should
be investigated in future studies. Accounting for an orthotropic
behavior would acquire more parameters to be estimated, which would
thus require more input data.

@@

% The choice of active model for the myocardium is another topic that
% should be addressed in more detail in future studies. In the author's
% opinion, the output of the two fundamentally different approaches may
% vary a lot. In particular, it is evident that the active strain
% formulation do a better job in fitting strain data. One
% hypothesis is that the amount of transverse active stresses should be
% adjusted to each individual. These transverse active stresses are
% naturally embedded in the active strain approach via a volume
% preserving active deformation.

Finally, when estimating high dimensional parameters, a natural question
concerning uniqueness of these estimates arises. It is obvious that if one allows for
enough degree of freedom in the parameter space, then it is possible
to fit almost any type of data. Therefore \emph{more work on ensuring
identifiability} of these estimates (Section
\ref{sec:intro_identifiability}) should be done. If the output of such
model should have any clinical utility, then uniqueness of the
estimated parameters is absolutely pivotal. Moreover, \emph{validation} of
these model is what matter most in terms of translating such computational
models into the clinic.
% Here we start by listing a couple of statements
% which should be taken into considerations.

### 2017-09-13 pre_thesis/list_of_papers.tex
Update to latest version

\textbf{Paper 1}
\hfill  \\ \noindent
Balaban, Gabriel, Henrik Finsberg, Hans Henrik Odland, Marie
E. Rognes, Stian Ross, Joakim Sundnes, and Samuel
Wall. \emph{High-resolution data assimilation of cardiac mechanics
  applied to a dyssynchronous ventricle.}  International journal for
numerical methods in biomedical engineering (2017). 
\\\\
\noindent
\textbf{Paper 2}
\hfill  \\ 
Finsberg, Henrik, Gabriel Balaban,  Stian Ross, Trine F. H\r{a}land,
Hans Henrik Odland, Joakim Sundnes, and Samuel
Wall. \emph{Estimating cardiac contraction through high resolution
  data assimilation of a personalized mechanical model.} Journal of
Computational Science (2017). doi:10.1016/j.jocs.2017.07.013

>>>>

\textbf{Paper 1}
\hfill  \\ \noindent
Balaban, Gabriel, Henrik Finsberg, Hans Henrik Odland, Marie
E. Rognes, Stian Ross, Joakim Sundnes, and Samuel
Wall. \emph{High-resolution data assimilation of cardiac mechanics
  applied to a dyssynchronous ventricle.}  International journal for
numerical methods in biomedical engineering
(2017). doi:10.1002/cnm.2863
Status: published
\\\\
\noindent
\textbf{Paper 2} 
\hfill  \\ 
Henrik, Finsberg, Gabriel Balaban,  Stian Ross, Trine F. H\r{a}land,
Hans Henrik Odland, Joakim Sundnes, and Samuel
Wall. \emph{Estimating cardiac contraction through high resolution
  data assimilation of a personalized mechanical model.} Journal of
Computational Science (2017). doi:10.1016/j.jocs.2017.07.013
\\Status: published
\\\\
\noindent
\textbf{Paper 3}
\hfill  \\
Henrik Finsberg, Ce Xi,  Ju Le Tan,  Liang Zhong, Martin Genet, Joakim
Sundnes, Lik Chuan Lee, and Samuel T. Wall.
\emph{Personalized bi-ventricular mechanical analysis using
  gradient-based optimization.} International journal for
numerical methods in biomedical engineering
\\Status: submitted
\\\\
\noident
\textbf{Paper 4}
\hfill  \\
Henrik Finsberg, John Aalen,Camilla K.
Larsen, Joakim Sundnes, Otto A. Smiseth, and 
Samuel T. Wall.
\emph{Assessment of regional myocardial work via adjoint-based
  data assimilation.} Biomechanics and Modeling in Mechanobiology
\\Status: submitted
